### 2018-03-31 paper/sexSelectionSpacing-ver1.tex
Edit Estimation Strategy

The unit of analysis is individual spells, the period from one birth, or marriage for
the first spell, to the next.
For some spells, the duration is censored, either because the survey
took place before the birth that would end the spell or because the couple
has finished childbearing.
The standard approach to address the censoring of spells, which I follow here, is 
to use a hazard model.
To capture the effects of sex selection, I extend the standard hazard model 
to allow for multiple exit states and a non-proportional hazard specification.

First, I use a competing risk framework with two possible exit states, either a boy
or a girl is born, because access to prenatal sex determination means that
the sex of children born is no longer necessarily a random event.
Prior research on birth spacing has used hazard models where the only 
possible exit state was the birth of a child. 
Having only one exit state works when there is no sex selection---making the 
sex of the child a random event---but not here because at least some couples use 
sex selection.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Second, I use a non-proportional hazard specification because
sex composition and the use of sex selection are likely to affect 
the shape of the hazard functions across groups.
Proportional hazard models, where covariates have a multiplicative 
effect on the hazard rate, are more efficient than non-proportional 
models, but only provided that the proportionality assumption hold.
If the proportionality assumption does not hold, the estimates
are biased.
It is unlikely---even in the absence of prenatal sex determination---that 
the effect of, for example, the sex composition of previous births have 
the same effect throughout the entire spell.
Assuming that the effect of sex composition is the same throughout a spell
is especially problematic for higher-order spells where different
sex composition of previous births can lead to substantial difference 
both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.
The introduction of prenatal sex determination is likely to exacerbate 
any bias from the proportionality assumption.
First, sex-selective abortions affect birth spacing, and use
varies across groups.
Second, the use of sex selection may vary within a spell, which means 
that the effects of covariates vary within the spell as well.
The use of a non-proportional specification also mitigates any potential 
effects of unobserved heterogeneity when used in conjunction with a 
flexible baseline hazard \citep{Dolton1995}.

In summary, my proposed model is a discrete time, non-proportional, competing risk 
hazard model with two exit states: either a boy or a girl is born.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

The hazard rate is specified as
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
where $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.
The explanatory variables in $\mathbf{Z}$ and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
This allows the effects of the main explanatory variables on the probabilities 
of having a boy, a girl, or no birth to vary over time within a spell.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.
% The exact specifications and the individual variables are described below.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is a period---here equal
to three months---the model can be estimated using a standard multinomial logit model.
if the data are transformed so the unit of analysis is a time period, in 
this case 3 months periods, rather than the individual woman, the model can be 
estimated using a standard multinomial logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is 0 if the
woman does not have a child in a given period (the base outcome), 1 if 
she gives birth to a son in that period, and 2 if she gives birth to 
a daughter in that period.

@@

An important issue is that the probability of ever having a next birth varies 
across groups.
For example, as the parity progression literature shows, couples with stronger 
son preference are less likely, for a given number of prior births, to ever have 
a next birth if they already have one or more sons.
Direct comparison of standard survival curves, therefore, tells us little 
about how the spread of sex selection affects birth spacing across, for
example, different sex compositions of prior children.

To overcome this problem, I condition on the likelihood of parity progression 
when examining birth spacing.
If, for example, women with a given set of characteristics have an 80\%
chance of ending a spell with a birth their median spell duration is when
their survival curve shows that 40\% are predicted to have exited the spell.
It is important to note that this is not the same as simply calculating 
the median \emph{observed} spell lengths among women who have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.

Median durations can then be averaged across women using the probabilities 
of ending a spell with a child as weights.
How well this approach works depends on whether the spell length covered is 
sufficiently long that few women are be likely to give birth after
the spell cut-off.
I discuss the choice of spell length below.
In addition to calculating conditional spell length for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Finally, with the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting across all $t$ 
gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.

>>>>

The unit of analysis is a spell, the period from one birth, or marriage for
the first spell, to the next.
For some spells, the duration is censored, for example, because the survey
took place before the birth that would end the spell or because the couple
has finished childbearing.
The standard approach to address the censoring of spells, which I follow here, is 
to use a hazard model.
To capture the effects of sex selection, I extend the standard hazard model 
to allow for multiple exit states and a non-proportional hazard specification.

First, I use a competing risk framework with two possible exit states, either a boy
or a girl is born, because access to prenatal sex determination means that
parents can choose the sex of children born.
Prior research on birth spacing, in contrast, relied on single
exit state hazard models.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}
Having only one exit state works when there is no sex selection---making the 
sex of the child a random event---but not here since at least some couples use 
sex selection.

Second, I use a non-proportional hazard specification because
sex composition and the use of sex selection are likely to affect 
the shape of the hazard functions across groups.
The use of a non-proportional specification also mitigates any potential 
effects of unobserved heterogeneity when used in conjunction with a 
flexible baseline hazard \citep{Dolton1995}.
Proportional hazard models, where covariates have a multiplicative 
effect on the hazard rate, are more efficient than non-proportional 
models, but only provided that the proportionality assumption hold.
If the proportionality assumption does not hold, the estimates
are biased.

It is unlikely---even in the absence of prenatal sex determination---that 
the effect of, for example, the sex composition of previous births have 
the same effect throughout the entire spell.
Assuming that the effect of sex composition is the same throughout a spell
is especially problematic for higher-order spells where different
sex composition of previous births can lead to substantial differences 
both in the likelihood of progressing to the next birth and
how soon couples want their next child if they are going to have one.

The introduction of prenatal sex determination is likely to exacerbate 
any bias from the proportionality assumption.
First, sex-selective abortions affect birth spacing, and use
varies across groups.
Second, the use of sex selection may vary within a spell, which means 
that the effects of covariates vary within the spell as well.

In summary, my proposed model is a discrete time, non-proportional, competing risk 
hazard model with two exit states: either a boy or a girl is born.
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$, and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: the birth of a boy, $j=1$, or the birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

The hazard rate is specified as
\begin{equation}
 h_{ijt} = \frac{\exp(D_j(t) + \alpha_{jt}'\mathbf{Z}_{it} + \beta_j'\mathbf{X}_{i})} 
 {1 + \sum_{l=1}^2 \exp(D_j(t) + \alpha_{lt}'\mathbf{Z}_{it} + \beta_l'\mathbf{X}_{i})} \: \: \; \; \;  j = 1,2
 \label{eq:hazard}
\end{equation}
where $D_{j}(t)$ is the piece-wise linear baseline hazard for outcome $j$, captured
by dummies and the associated coefficients,
\begin{equation}
D_j(t) = \gamma_{j1} D_1 + \gamma_{j2} D_2 + \ldots + \gamma_{jT} D_T,
\end{equation}
where $D_m = 1$ if $t=m$ and zero otherwise.
This approach to modeling the baseline hazard is flexible and does not place 
overly strong restrictions on the baseline hazard.

The explanatory variables in $\mathbf{Z}$ and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The non-proportionality allows the effects of the main explanatory variables on the probabilities 
of having a boy, a girl, or no birth to vary over time within a spell.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.
% The exact specifications and the individual variables are described below.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, transforming the data, so each observation is an interval---here equal
to three months---the model can be estimated using a standard multinomial logit model.%
\footnote{
A potential issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
I, therefore, include a proxy for fecundity discussed in Section \ref{sec:data}.
Also, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are substantially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is 0 if the
woman does not have a child in a given interval (the base outcome), 1 if 
she gives birth to a son in that interval, and 2 if she gives birth to 
a daughter in that interval.

@@

An important issue is that the probability of ever having a next birth varies 
across groups.
For example, as the parity progression literature shows, couples with stronger 
son preference are less likely, for a given number of prior births, to ever have 
a next birth if they already have one or more sons.
Direct comparison of standard survival curves, therefore, tells us little 
about how the spread of sex selection affects birth spacing across groups.

To overcome this problem, I condition on the predicted likelihood of parity 
progression when examining birth spacing measures, such as the median duration 
to a birth.
How well this approach works depends on whether the spell length covered is 
sufficiently long that few women are likely to give birth after the spell cut-off.
I discuss the choice of spell length below.
It is important to note that the approach is not the same as merely calculating 
the birth spacing measures for women who already have a given 
parity child in the survey because that number does not take into account
the censoring of spells that will eventually lead to a birth.
In addition to calculating conditional birth spacing measures for different 
percentages, I also present graphs of both standard survival curves and
survival curves conditional on parity progression (which therefore
begin at 100\% and end at 0\%).

Finally, given the predicted probabilities of having a boy and of having 
a girl for each $t$ within a spell, is it easy to calculate the estimated 
percentage of children born that are boys, $\hat{Y}$, at each $t$  
\begin{equation}
\hat{Y}_t 
= 
\frac{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )}
{ P(b_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t) + P(g_{t} | \mathbf{X}_{k}, \mathbf{Z}_{kt},t )} 
\times 100.
\label{eq:probability_son}
\end{equation}
Combining the percentage boys and the likelihood of exiting the spell 
across all $t$ gives the predicted percent boys born over the entire spell.
In addition to the predicted percent boys over the entire spell, 
I also present graphs of how the percentage boys born vary across time
within a spell, together with the associated confidence interval for given 
values of explanatory variables calculated using the Delta method.
